Classless Java

### 2016-03-23 OOPSLA16/sections/Introduction.tex
references

Object oriented languages strive to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas have emerged to this end: prototype
based (PB)~\cite{} and class based (CB) languages~\cite{}.  In prototype based
languages objects inherit from other objects, and thus objects own
both behaviour and state (and objects is all you have).
In class based languages an object is instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

@@

In the presence of subtyping, abstract state operations often require
special care, as their types need to be refined. Object interfaces
provide support for type-refinement and can automatically produce code
that deals with type-refinement adequately. In contrast, typical CB
languages, such as Java or Scala, require verbose explicit
type-refinement. We believe that such verbosity hindered and slowed
down the discovery of useful programming pattens involving
type-refinement. In particular, a recently discovered
solution~\cite{trivialsolutiontotheEP} to the Expression
Problem~\cite{} in Java-like languages, shows how easy it is to solve
the problem using only type-refinement. However it took nearly 20
years since the formulation of the problem for that
solution to be presented in the literature. In IB programming, due to
automatic type-refinement, that solution should have been obvious from
day one.

One important advantage of the use of abstract state operations and
automatic type-refinement is that it offers a new form of
\emph{type-safe covariant mutable state}. That is, in IB programming,
it is possible to refine \emph{mutable} ``fields'' in subtypes. This is
typically forbidden in CB programming languages (including Java and
Scala), since it is widely known that \emph{naive} type-refinement of
mutable fields is not type-safe. While covariant refinement of mutable
fields is supported by some type systems~\cite{}, this requires
significant complexity and restrictions to ensure that all uses of
covariant state are indeed type-safe.

@@

Finally, another advantage of IB programming is its support for
multiple inheritance. The literature provides good examples on how
easy and modular it is to combine multiple sources of pure behaviour,
using mechanisms such as traits~\cite{scharli03traits}. In Java
multiple \emph{interface} inheritance has been supported since
inception, and in Java8 default methods~\cite{} bring some of the
advantages of traits into Java. In contrast, the literature~\cite{}
is also rich on how hard it is to modularly combine multiple sources
of behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes. In IB programming there is only multiple
interface inheritance, yet programmers can still use state via the
abstract state operations. IB programming enables powerful 
idioms using multiple inheritance and state.

@@

While the Java embedding has obvious advantages from the practical
point-of-view, it also imposes some limitations that a new IB language
would not have. In particular, supporting proper encapsulation is
difficult in Java due to limitations of Java interfaces. We discuss
these limitations, possible workarounds, and native language support in Section~\ref{}.

>>>>

Object oriented languages strive to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas have emerged to this end: prototype
based (PB)~\cite{Ungar87self} and class based (CB) languages such as
Java or Scala.  In prototype based
languages objects inherit from other objects, and thus objects own
both behaviour and state (and objects is all you have).
In class based languages an object is instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

@@

In the presence of subtyping, abstract state operations often require
special care, as their types need to be refined. Object interfaces
provide support for type-refinement and can automatically produce code
that deals with type-refinement adequately. In contrast, typical CB
languages, such as Java or Scala, require verbose explicit
type-refinement. We believe that such verbosity hindered and slowed
down the discovery of useful programming pattens involving
type-refinement. In particular, a recently discovered
solution~\cite{eptrivially} to the Expression
Problem~\cite{wadler98expression} in Java-like languages, shows how easy it is to solve
the problem using only type-refinement. However it took nearly 20
years since the formulation of the problem for that
solution to be presented in the literature. In IB programming, due to
automatic type-refinement, that solution should have been obvious from
day one.

One important advantage of the use of abstract state operations and
automatic type-refinement is that it offers a new form of
\emph{type-safe covariant mutable state}. That is, in IB programming,
it is possible to refine \emph{mutable} ``fields'' in subtypes. This is
typically forbidden in CB programming languages (including Java and
Scala), since it is widely known that \emph{naive} type-refinement of
mutable fields is not type-safe. While covariant refinement of mutable
fields is supported by some type systems~\cite{bruce98astatically,McDirmid01jiazzi,ernst01family,ernst06virtual}, this requires
significant complexity and restrictions to ensure that all uses of
covariant state are indeed type-safe.

@@

Finally, another advantage of IB programming is its support for
multiple inheritance. The literature provides good examples on how
easy and modular it is to combine multiple sources of pure behaviour,
using mechanisms such as traits~\cite{scharli03traits}. In Java
multiple \emph{interface} inheritance has been supported since
inception, and in Java8 default methods~\cite{goetz12fdefenders} bring some of the
advantages of traits into Java. In contrast, the literature~\cite{Sak89dis,bracha90mixin,malayeri2009cz}
is also rich on how hard it is to modularly combine multiple sources
of behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes. In IB programming there is only multiple
interface inheritance, yet programmers can still use state via the
abstract state operations. IB programming enables powerful 
idioms using multiple inheritance and state.

@@

While the Java embedding has obvious advantages from the practical
point-of-view, it also imposes some limitations that a new IB language
would not have. In particular, supporting proper encapsulation is
difficult in Java due to limitations of Java interfaces. We discuss
these limitations, possible workarounds, and native language support in Section~\ref{sec:futurework}.
